Estimating duration in the presence of misclassification

### 2024-08-02 biology-data.tex
corrections I spotted on paper

The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
ATACCC enrolled a total of 738 contacts across two waves.
The first wave enrolled 393 contacts between 13 Sep 2020 and 31 Mar 2021, when pre-Alpha and Alpha variants were circulating.
The second wave enrolled 345 contacts between 24 May 2021 and 28 Oct 2021, when the Delta variant was circulating.
Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
Other data such as demography and symptoms were also collected.
For further details see \textcite{singanayagamDuration,hakkiOnset}.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
Denote these by $p^{(j)}_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
Define a (sub) population of interest as a set of strata $\set{S}$.
Then the poststratified prevalence in the (sub)population on day $t$ is
\begin{align}
p^{(j)}_t(\set{S})
&= \frac{\sum_{i \in \set{S}} p^{(j)}_{t,i} N_i}{\sum_{i \in \set{S}} N_i}
\end{align}
where $N_i$ is the population size of stratum $i$.
The population numbers for each stratum are provided by the ONS.

>>>>

The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
ATACCC enrolled a total of 738 contacts across two waves.
The first wave enrolled 393 contacts between 13 Sep 2020 and 31 Mar 2021, when pre-Alpha and Alpha variants were circulating.
The second wave enrolled 345 contacts between 24 May 2021 and 28 Oct 2021, when the Delta variant was circulating.
Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
Other data such as demography and symptoms were also collected.
For further details see \textcite{singanayagamCommunity,hakkiOnset}.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
Denote these by $p^{(j)}_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
Define a (sub) population of interest as a set of strata $\set{S}$.
Then the $j$th posterior sample of poststratified prevalence in the (sub)population on day $t$ is
\begin{align}
p^{(j)}_t(\set{S})
&= \frac{\sum_{i \in \set{S}} p^{(j)}_{t,i} N_i}{\sum_{i \in \set{S}} N_i}
\end{align}
where $N_i$ is the population size of stratum $i$.
The population numbers for each stratum are provided by the ONS.

### 2024-08-02 incidence-prevalence.tex
corrections I spotted on paper

The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.

@@

Denote by $R_{t,t'}$ the number of infection episodes with $B_i=t$ and $E_i=t'$.
For any infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
As I assume these durations are independent~\autocite{paganoHIV}:
\begin{align}
\begin{bmatrix}
  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
\end{bmatrix} \mid Z_t
\sim \MN \left(
  Z_t, 
  \begin{bmatrix}
    f_D(1) \\ f_D(2) \\ \vdots \\ f_D(d_\text{max})
  \end{bmatrix}
\right).
\end{align}
Now, denote by $P_{t_z,t_p}$ the number of individuals infected at $t_z$ and prevalent at $t_p$.
Formally, $P_{t_z,t_p}$ is the infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
$P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
Hence:
\begin{align}
    P_{t_z,t_p} = \begin{cases}
      0 &t_p < t_z\\
      Z_{t_z} - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
      0 &t_p \geq t_z + d_\text{max}.
  \end{cases} \label{inc-prev:eq:Ptt-to-R}
\end{align}
Finally, express $P_t$ as the total number of people prevalent at time $t$, regardless of when they were infected.
The zeroes in \cref{inc-prev:eq:Ptt-to-R} determine the sum's limits.
\begin{align}
  P_t
  &= \sum_{t_z=-\infty}^\infty P_{t_z,t} \\
  &= \sum_{i=0}^{\dmax-1} P_{t-i,t} \label{inc-prev:eq:Pt-to-Ptt} \\
  &= \sum_{i=0}^{\dmax-1} \left(Z_{t_z} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) &\text{by \cref{inc-prev:eq:Ptt-to-R}}\label{inc-prev:eq:Pt-to-Rtt}.
\end{align}

>>>>

The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
Ignoring the latent period will again underestimate the reproduction number~\autocite{wearingAppropriate}.
The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.

@@

Denote by $R_{t,t'}$ the number of infection episodes with $B_i=t$ and $E_i=t'$.
For any infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
As I assume these durations are independent~\autocite{paganoHIV}:
\begin{align}
\begin{bmatrix}
  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
\end{bmatrix} \mid Z_t
\sim \MN \left(
  Z_t, 
  \begin{bmatrix}
    f_D(1) \\ f_D(2) \\ \vdots \\ f_D(d_\text{max})
  \end{bmatrix}
\right).
\end{align}
Now, denote by $P_{t_z,t_p}$ the number of individuals infected at $t_z$ and prevalent at $t_p$.
Formally, $P_{t_z,t_p}$ is the number of infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
$P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
Hence:
\begin{align}
    P_{t_z,t_p} = \begin{cases}
      0 &t_p < t_z\\
      Z_{t_z} - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
      0 &t_p \geq t_z + d_\text{max}.
  \end{cases} \label{inc-prev:eq:Ptt-to-R}
\end{align}
Finally, express $P_t$ as the total number of people prevalent at time $t$, regardless of when they were infected.
The zeroes in \cref{inc-prev:eq:Ptt-to-R} determine the sum's limits.
\begin{align}
  P_t
  &= \sum_{t_z=-\infty}^\infty P_{t_z,t} \\
  &= \sum_{i=0}^{\dmax-1} P_{t-i,t} \label{inc-prev:eq:Pt-to-Ptt} \\
  &= \sum_{i=0}^{\dmax-1} \left(Z_{t_z} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) &\text{by \cref{inc-prev:eq:Ptt-to-R}}\label{inc-prev:eq:Pt-to-Rtt}.
\end{align}

### 2024-08-02 introduction.tex
corrections I spotted on paper

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf (probability density function) of a distribution, or a transformation of it.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys which aimed to recruit representative samples of the general population; such surveys are considered the gold-standard for conducting population surveillance~\autocite{mccabeCISincidence}.
To achieve this aim, they invited individuals selected at random from the population; however, not all individuals selected responded to the invite introducing some non-representativeness (see \cref{E-biology-data:sec:MRP}).
These designs aim to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

>>>>

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on suffering the consequence of interest, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf (probability density function) of a distribution, or a transformation of it.

@@

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys which aimed to recruit representative samples of the general population; such surveys are considered the gold-standard for conducting population surveillance~\autocite{mccabeCISincidence}.
To achieve this aim, they invited individuals selected at random from the population; however, not all individuals selected responded to the invite introducing some non-representativeness (see \cref{E-biology-data:sec:MRP}).
These designs aim to avoid the biases of community testing and observe the general population.
These surveys use RT-PCR (Reverse Transcription Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
